Statistical Machine Learning Models on A Bank Marketing dataset

### 2025-12-07 Slides.Rmd
Updated Slides KNN and boosted trees

## KNN 1/3: Feature Preparation 
* **Discuss:** **KNN** requires scaling: all numeric predictors were **standardized** (mean=0, SD=1) to prevent features with larger scales from dominating the distance calculation.
* **Discuss:** The non-parametric nature means **no explicit feature selection** via coefficients, but the scaling step is vital for ensuring equal contribution from all dimensions.
* **Image/Plot Idea:** A visual comparison: a scatter plot of two features *unscaled* vs. the same features *scaled/standardized*.

## KNN 2/3: Hyperparameter Tuning 
* **Discuss:** The sole hyperparameter, **$k$** (the number of neighbors), determines the model's complexity (i.e., the bias-variance trade-off).
* **Discuss:** **5-Fold Cross-Validation** was performed across a range of $k$ values to find the optimal $k$ that minimized the mean CV error.
* **Image/Plot Idea:** A plot showing **CV Test Error vs. $k$**, illustrating the U-shape curve where the optimal $k$ is selected.

@@

@@

## Boosting 1/3: Model Formulation
* **Discuss:** **Boosting** (e.g., Gradient Boosting) is an ensemble method that sequentially builds weak learners (trees) to minimize the classification error of the previous step.
* **Discuss:** This model is expected to be our **highest-performing benchmark** but sacrifices interpretability due to its complex, additive structure.
* **Image/Plot Idea:** A conceptual diagram illustrating the boosting process (weak learners combining sequentially to form a strong learner).

## Boosting 2/3: Hyperparameter Tuning 
* **Discuss:** Multiple hyperparameters were tuned using **Cross-Validation**, including the **number of trees** (M), **interaction depth** (complexity of weak learners), and **learning rate** ($\eta$).
* **Discuss:** The tuning process involves finding the right balance between these parameters to ensure the model generalizes well without overfitting the training data.
* **Image/Plot Idea:** A plot showing the **Error Rate vs. Number of Trees**, highlighting where the Test Error flattens or begins to increase (overfitting).

## Boosting 3/3: Evaluation & Importance 
* **Discuss:** The final Boosting model was evaluated on the **Test Set**. Confirm that it achieved the **highest AUC**, setting the benchmark for predictive accuracy.
* **Discuss:** While coefficients are not available, discuss the model's **Variable Importance Plot** as the only glimpse into its internal workings.
* **Image/Plot Idea:** The **Variable Importance Plot** from the Boosting model, which should confirm that the most important features align reasonably well with LogReg's significant coefficients.

>>>>

## KNN 1/3: Feature Preparation 
* **Discuss:** **KNN** requires scaling: all numeric predictors were **standardized** (mean=0, SD=1) to prevent features with larger scales from dominating the distance calculation.
* **Discuss:** The non-parametric nature means **no explicit feature selection** via coefficients, but the scaling step is vital for ensuring equal contribution from all dimensions.
* **Image/Plot Idea:** A visual comparison: a scatter plot of two features *unscaled* vs. the same features *scaled/standardized*. 
```{r}
library(dplyr)
library(class)
library(caret)

set.seed(123)
train_index <- sample(nrow(bank_full), 0.8 * nrow(bank_full))
bank_train <- bank_full[train_index, ]
bank_test  <- bank_full[-train_index, ]

numeric_vars <- names(bank_full)[sapply(bank_full, is.numeric)]
numeric_vars <- setdiff(numeric_vars, "y")

train_x <- bank_train[, numeric_vars, drop = FALSE]
test_x  <- bank_test[, numeric_vars, drop = FALSE]

train_x <- bank_train[, numeric_vars]
test_x  <- bank_test[, numeric_vars]

train_x_scaled <- scale(train_x)
test_x_scaled <- scale(
  test_x,
  center = attr(train_x_scaled, "scaled:center"),
  scale  = attr(train_x_scaled, "scaled:scale")
)

train_y <- bank_train$y
test_y  <- bank_test$y

best_k <- 5

par(mfrow = c(1, 2))

plot(
  bank_full$age, bank_full$balance,
  xlab = "Age", ylab = "Balance",
  main = "Unscaled Features"
)

scaled_two <- scale(bank_full[, c("age", "balance")])

plot(
  scaled_two[, "age"], scaled_two[, "balance"],
  xlab = "Scaled Age", ylab = "Scaled Balance",
  main = "Scaled Features (mean = 0, sd = 1)"
)

par(mfrow = c(1, 1))

i <- 1  
test_point <- test_x_scaled[i, , drop = FALSE]

distances <- apply(
  train_x_scaled,
  1,
  function(row) sqrt(sum((row - test_point)^2))
)

nn_idx <- order(distances)[1:best_k]

knn_demo_pred <- knn(
  train = train_x_scaled,
  test  = test_point,
  cl    = train_y,
  k     = best_k
)

cat("Test customer index:", i, "\n")
cat("KNN predicted class:", knn_demo_pred, "\n")
cat("True class:", test_y[i], "\n\n")

cat("Neighbor outcome breakdown:\n")
print(table(train_y[nn_idx]))

cat("\nNearest neighbor details:\n")
print(head(bank_train[nn_idx, c("age", "balance", "duration", "campaign", "y")]))

```
```{r}
library(class)

numeric_vars <- names(bank_full)[sapply(bank_full, is.numeric)]
numeric_vars <- setdiff(numeric_vars, "y")

train_x <- bank_train[, numeric_vars, drop = FALSE]
test_x  <- bank_test[, numeric_vars, drop = FALSE]

train_x_scaled <- scale(train_x)
test_x_scaled <- scale(
  test_x,
  center = attr(train_x_scaled, "scaled:center"),
  scale  = attr(train_x_scaled, "scaled:scale")
)

train_y <- bank_train$y
test_y  <- bank_test$y

knn_pred <- knn(
  train = train_x_scaled,
  test  = test_x_scaled,
  cl    = train_y,
  k     = 5
)

knn_accuracy <- mean(knn_pred == test_y)
knn_error <- 1 - knn_accuracy
cat("KNN Test Accuracy:", round(knn_accuracy * 100, 2), "%\n")
cat("KNN Test Error:", round(knn_error * 100, 2), "%\n")

```
```

## KNN 2/3: Hyperparameter Tuning 
* **Discuss:** The sole hyperparameter, **$k$** (the number of neighbors), determines the model's complexity (i.e., the bias-variance trade-off).
* **Discuss:** **5-Fold Cross-Validation** was performed across a range of $k$ values to find the optimal $k$ that minimized the mean CV error.
* **Image/Plot Idea:** A plot showing **CV Test Error vs. $k$**, illustrating the U-shape curve where the optimal $k$ is selected.
```{r}
library(class)
library(caret)

set.seed(123)

k_values <- 1:25                  
cv_error <- numeric(length(k_values))

folds <- createFolds(train_y, k = 5, returnTrain = FALSE)

for (i in seq_along(k_values)) {
  k <- k_values[i]
  fold_err <- numeric(length(folds))

  for (j in seq_along(folds)) {
    test_idx  <- folds[[j]]
    train_idx <- setdiff(seq_along(train_y), test_idx)

    fold_train_x <- train_x_scaled[train_idx, , drop = FALSE]
    fold_test_x  <- train_x_scaled[test_idx,  , drop = FALSE]
    fold_train_y <- train_y[train_idx]
    fold_test_y  <- train_y[test_idx]

    pred <- knn(
      train = fold_train_x,
      test  = fold_test_x,
      cl    = fold_train_y,
      k     = k
    )

    fold_err[j] <- mean(pred != fold_test_y)
  }

  cv_error[i] <- mean(fold_err)
}

best_k <- k_values[which.min(cv_error)]
best_k
plot(
  k_values, cv_error,
  type = "b",
  xlab = "Number of neighbors (k)",
  ylab = "5-fold CV Test Error",
  main = "KNN: 5-fold CV Test Error vs k"
)

abline(v = best_k, lty = 2)

@@

```{r}
library(pROC)
library(caret)

set.seed(123)

knn_final <- knn(
  train = train_x_scaled,
  test  = test_x_scaled,
  cl    = train_y,
  k     = best_k,
  prob  = TRUE
)

knn_prob_attr <- attr(knn_final, "prob")
knn_prob_yes <- ifelse(knn_final == "yes", knn_prob_attr, 1 - knn_prob_attr)

test_y_fac <- factor(test_y, levels = c("no", "yes"))

knn_roc  <- roc(response = test_y_fac, predictor = knn_prob_yes, levels = c("no", "yes"))
knn_auc  <- as.numeric(auc(knn_roc))

knn_cm      <- confusionMatrix(knn_final, test_y_fac, positive = "yes")
knn_recall  <- knn_cm$byClass["Recall"]

logit_model <- glm(y ~ ., data = bank_train, family = binomial)

logit_prob <- predict(logit_model, newdata = bank_test, type = "response")
logit_pred <- ifelse(logit_prob >= 0.5, "yes", "no")
logit_pred <- factor(logit_pred, levels = c("no", "yes"))

logit_roc <- roc(response = test_y_fac, predictor = logit_prob, levels = c("no", "yes"))
logit_auc <- as.numeric(auc(logit_roc))

logit_cm     <- confusionMatrix(logit_pred, test_y_fac, positive = "yes")
logit_recall <- logit_cm$byClass["Recall"]

results <- data.frame(
  Model  = c("Logistic Regression", paste0("KNN (k = ", best_k, ")")),
  AUC    = c(logit_auc, knn_auc),
  Recall = c(as.numeric(logit_recall), as.numeric(knn_recall))
)

results$AUC    <- round(results$AUC, 3)
results$Recall <- round(results$Recall, 3)

results

@@

## Boosting 1/3: Model Formulation
* **Discuss:** **Boosting** (e.g., Gradient Boosting) is an ensemble method that sequentially builds weak learners (trees) to minimize the classification error of the previous step.
* **Discuss:** This model is expected to be our **highest-performing benchmark** but sacrifices interpretability due to its complex, additive structure.
* **Image/Plot Idea:** A conceptual diagram illustrating the boosting process (weak learners combining sequentially to form a strong learner). 
```{r}
library(gbm)

bank_train$y_num <- ifelse(bank_train$y == "yes", 1, 0)
bank_test$y_num  <- ifelse(bank_test$y  == "yes", 1, 0)

set.seed(123)

boost_model <- gbm(
  formula = y_num ~ . - y,       
  data = bank_train,
  distribution = "bernoulli",   
  n.trees = 1000,                
  interaction.depth = 3,         
  shrinkage = 0.01,             
  n.minobsinnode = 10,
  verbose = FALSE
)

boost_prob <- predict(
  boost_model,
  newdata = bank_test,
  n.trees = 1000,
  type = "response"
)

boost_pred <- ifelse(boost_prob > 0.5, "yes", "no")

boost_accuracy <- mean(boost_pred == bank_test$y)
boost_error    <- 1 - boost_accuracy

cat("Boosting Test Accuracy:", round(boost_accuracy * 100, 2), "%\n")
cat("Boosting Test Error:",   round(boost_error * 100, 2), "%\n")

plot(
  boost_model$train.error,
  type = "l",
  xlab = "Number of Trees (Iterations)",
  ylab = "Training Loss",
  main = "Boosting: Sequential Error Reduction"
)

## Boosting 2/3: Hyperparameter Tuning 
* **Discuss:** Multiple hyperparameters were tuned using **Cross-Validation**, including the **number of trees** (M), **interaction depth** (complexity of weak learners), and **learning rate** ($\eta$).
* **Discuss:** The tuning process involves finding the right balance between these parameters to ensure the model generalizes well without overfitting the training data.
* **Image/Plot Idea:** A plot showing the **Error Rate vs. Number of Trees**, highlighting where the Test Error flattens or begins to increase (overfitting).
```{r}
boost_model <- gbm(
  formula = y_num ~ . - y,
  data = bank_train,
  distribution = "bernoulli",
  n.trees = 1000,
  interaction.depth = 3,
  shrinkage = 0.01,
  n.minobsinnode = 10,
  verbose = FALSE
)

M_values <- seq(50, 1000, by = 50)
test_error <- numeric(length(M_values))

for (i in seq_along(M_values)) {
  M <- M_values[i]

  prob_M <- predict(
    boost_model,
    newdata = bank_test,
    n.trees = M,
    type = "response"
  )

  pred_M <- ifelse(prob_M > 0.5, "yes", "no")
  test_error[i] <- mean(pred_M != bank_test$y)
}

best_idx <- which.min(test_error)
best_M   <- M_values[best_idx]

cat("Best number of trees based on Test Error:", best_M, "\n")

plot(
  M_values, test_error,
  type = "b",
  xlab = "Number of Trees (M)",
  ylab = "Test Error",
  main = "Boosting: Test Error vs Number of Trees"
)

abline(v = best_M, lty = 2)

## Boosting 3/3: Evaluation & Importance 
* **Discuss:** The final Boosting model was evaluated on the **Test Set**. Confirm that it achieved the **highest AUC**, setting the benchmark for predictive accuracy.
* **Discuss:** While coefficients are not available, discuss the model's **Variable Importance Plot** as the only glimpse into its internal workings.
* **Image/Plot Idea:** The **Variable Importance Plot** from the Boosting model, which should confirm that the most important features align reasonably well with LogReg's significant coefficients.
```{r}
library(pROC)

boost_prob <- predict(
  boost_model,
  newdata = bank_test,
  n.trees = 1000,       
  type = "response"
)

test_y_fac <- factor(bank_test$y, levels = c("no", "yes"))

boost_roc <- roc(
  response  = test_y_fac,
  predictor = boost_prob,
  levels    = c("no", "yes")
)

boost_auc <- as.numeric(auc(boost_roc))
cat("Boosting Test AUC:", round(boost_auc, 3), "\n")

vi <- summary(boost_model, plotit = FALSE)

top_n <- min(10, nrow(vi))
vi_top <- vi[1:top_n, ]

barplot(
  height = vi_top$rel.inf,
  names.arg = vi_top$var,
  las = 2,
  ylab = "Relative Influence",
  main = "Boosting: Variable Importance (Top Features)"
)
